---
title: "Dataset Report - Group 2"
subtitle: "COVID-19 Mortality & Structural Predictors"
author: "Emily Panzl & Max Elixhauser"
date: today
format:
  PrettyPDF-pdf:
    include-in-header:
      - text: |
          \usepackage{fvextra}
          \DefineVerbatimEnvironment{Highlighting}{Verbatim}{
            breaklines,
            breaksymbolleft={},
            commandchars=\\\{\}
          }
code-fold: true
echo: false
message: false
warning: false
---

## Introduction

This report presents the dataset selected for our multivariate analysis project. 
We investigate structural country-level predictors of COVID-19 mortality, using 
cumulative excess mortality as our outcome variable. The central question is: 
*what structural factors explain variation in COVID-19 mortality across countries?*

## Data Sources

We use two datasets from Our World in Data (OWID), both licensed under CC BY 4.0:

- **Excess Mortality** (HMD/WMD via OWID): cumulative excess deaths per million, 
  monthly, 126 countries, January 2020 - December 2023.  
  Source: https://catalog.ourworldindata.org/garden/excess_mortality/latest/excess_mortality/excess_mortality.csv

- **OWID Compact COVID-19 Dataset**: structural country-level variables (GDP, 
  demographics, health infrastructure).  
  Source: https://catalog.ourworldindata.org/garden/covid/latest/compact/compact.csv

## Variables

| Variable | Description | Source |
|---|---|---|
| `excess_mort` | Cumulative excess deaths per million (snapshot ~2023-05-05) | HMD/WMD via OWID |
| `gdp_per_capita` | GDP per capita, PPP (constant 2017 int'l $) | World Bank via OWID |
| `median_age` | Median population age (years) | UN WPP via OWID |
| `hospital_beds_per_thousand` | Hospital beds per 1,000 people | World Bank via OWID |

## Variable Guide - WTF is a:

**Excess Mortality** is not just confirmed COVID deaths. It measures: 
*reported all-cause deaths minus expected deaths based on 2015–2019 baseline*. 
If a country normally has 1,000 deaths/month and had 1,300 during COVID, 
excess = 300. Scaled per million for fair cross-country comparison.

Negative values mean *fewer* deaths than expected (e.g. strict lockdowns reducing accidents like car crashes).

**GDP per capita (PPP)** is the total economic output per person, adjusted for 
purchasing power (so $1 in Bolivia ≠ $1 in Germany). Higher = wealthier country (Note: high GDP per capita reflects national output, not necessarily that individual citizens are wealthy as income inequality matters too.)
Unit: constant 2017 international dollars.

**Median Age** the age that splits a population 50/50. 
Japan median age ~48 = half the population is older than 48. (welp.)
Niger median age ~15 = very young population. 
Higher median age = more elderly people at risk from COVID.

**Hospital Beds per 1,000** is number of inpatient hospital beds available 
per 1,000 residents. Proxy for healthcare infrastructure capacity. 
Note: this measures *capacity*, not quality or staffing.

**Pearson r** measures linear correlation between two variables, 
ranging from -1 to +1. 
r = +1: perfect positive relationship (both go up together). 
r = -1: perfect negative (one goes up, other goes down). 
r = 0: no linear relationship. 
Rule of thumb: |r| < 0.1 negligible, 0.1–0.3 weak, 0.3–0.5 moderate, > 0.5 strong.
(We use it here to get a first sense of which variables move together with excess mortality.. before building more complex models in the final project.)

## Data Preparation

```{r}
# install.packages(c("readr", "dplyr", "countrycode")) # for the install
library(readr)
library(dplyr)
library(countrycode)

# 1) Excess Mortality
excess_catalog <- read_csv("data/excess_mortality.csv", show_col_types = FALSE)

excess_snap <- excess_catalog |>
  # drop rows without excess mortality data
  filter(!is.na(cum_excess_per_million_proj_all_ages)) |>
  # countrycode converts country names to ISO3 codes (e.g. "Albania" -> "ALB")
  # custom_match handles edge cases that countrycode can't resolve automatically
  mutate(code = countrycode(entity, "country.name", "iso3c",
    custom_match = c("Kosovo" = "XKX"))) |>
  # drop rows where ISO mapping failed (e.g. regional aggregates)
  filter(!is.na(code)) |>
  # per country: pick the observation closest to 2023-05-05 (WHO end-of-emergency)
  group_by(code) |>
  slice_min(abs(date - as.Date("2023-05-05")), with_ties = FALSE) |>
  ungroup() |>
  select(code, excess_mort = cum_excess_per_million_proj_all_ages)

# 2) Structural Variables (OWID compact)
owid_struct <- read_csv("data/owid-covid-data.csv", show_col_types = FALSE) |>
  # OWID compact contains regional aggregates prefixed with "OWID_" - drop those
  filter(!startsWith(code, "OWID_")) |>
  # structural variables (GDP, age etc.) repeat on every row per country
  # just take the latest row to get one row per country
  group_by(code) |>
  slice_max(date, with_ties = FALSE) |>
  ungroup() |>
  select(code, gdp_per_capita, median_age,
         hospital_beds_per_thousand, continent)

# 3) Join & export
# inner_join: only keep countries present in both datasets
# complete.cases: drop any remaining rows with NAs across all variables
dat_analysis <- inner_join(excess_snap, owid_struct, by = "code") |>
  filter(complete.cases(pick(everything())))

# export for EDA
write_csv(dat_analysis, "data/dat_analysis.csv")
```

**n = `r nrow(dat_analysis)` countries** with complete data across all variables.


## What the Data Already Tells Us
```{r}
# quick summary
# descriptive summary all variables
dat_analysis |>
  select(excess_mort, gdp_per_capita, median_age, hospital_beds_per_thousand) |>
  summarise(across(everything(), list(
    mean = ~round(mean(.), 1),
    median = ~round(median(.), 1),
    min = ~round(min(.), 1),
    max = ~round(max(.), 1)
  ))) |>
  tidyr::pivot_longer(everything(),
    names_to = c("variable", "stat"),
    names_sep = "_(?=[^_]+$)") |>
  tidyr::pivot_wider(names_from = stat, values_from = value) |>
  knitr::kable(caption = "Descriptive summary of all variables (n = 99 countries)")
```


Table: Descriptive summary of all variables (n = 99 countries)

|variable                   |    mean|  median|    min|      max|
|:--------------------------|-------:|-------:|------:|--------:|
|excess_mort                |  3018.5|  2616.5| -654.3|   9767.1|
|gdp_per_capita             | 33034.6| 27450.4| 4137.3| 117747.0|
|median_age                 |    35.5|    35.8|   21.7|     48.5|
|hospital_beds_per_thousand |     3.7|     3.0|    0.4|     13.0|

Massive variation across all variables as GDP per capita from **$4,137** to **$117,747**, 
median age from **21.7** (sub-Saharan Africa vibes) to **48.5** years (Japan... o7), 
hospital beds from **0.4** to **13.0** per 1,000. And excess mortality from **`r round(min(dat_analysis$excess_mort))`** 
(some countries had *fewer* deaths than expected.. so strict lockdowns cutting car crashes etc.) 
to **`r round(max(dat_analysis$excess_mort))`** deaths per million.

That's a **`r round(max(dat_analysis$excess_mort) / abs(min(dat_analysis$excess_mort)))`$\times$ difference** between best and worst outcomes.. which begs the question: why though? Can I haz cheezburgers?

A first look at the predictors gives us a hint:
```{r}
# correlation with excess mortality
dat_analysis |>
  select(excess_mort, gdp_per_capita, median_age, hospital_beds_per_thousand) |>
  cor(use = "complete.obs", method = "pearson") |>
  round(2) |>
  knitr::kable(caption = "Pearson correlation matrix")
```

Table: Pearson correlation matrix

|                           | excess_mort| gdp_per_capita| median_age| hospital_beds_per_thousand|
|:--------------------------|-----------:|--------------:|----------:|--------------------------:|
|excess_mort                |        1.00|          -0.28|       0.27|                       0.21|
|gdp_per_capita             |       -0.28|           1.00|       0.45|                       0.10|
|median_age                 |        0.27|           0.45|       1.00|                       0.59|
|hospital_beds_per_thousand |        0.21|           0.10|       0.59|                       1.00|

Three things stand out already:

- **GDP** is *negatively* correlated with excess mortality (r = -0.28): richer countries 
  had fewer excess deaths on average.
- **Median age** is *positively* correlated (r = 0.27): older populations were hit harder.
- **Hospital beds** is *positively* correlated (r = 0.21): counterintuitive at first glance, 
  but notice that hospital beds and median age are strongly correlated with each other (r = 0.59). 
  Countries with older populations tend to have built more hospital capacity over time...
  so beds are partly picking up the age effect. Classic suppression. :>

## Exploratory Data Analysis

```{python}
import pandas as pd

dat = pd.read_csv("data/dat_analysis.csv")
dat.describe()
```

## Potential Research Questions

For the final project, we intend to investigate the following (subject to change after the April 16 pitch):

...

## References

- Human Mortality Database (2024); World Mortality Dataset (2024); Karlinsky & Kobak (2021) – processed by Our World in Data. CC BY 4.0.
- Our World in Data COVID-19 Compact Dataset. CC BY 4.0.
